Grappa: Scalable Computing on Mass-Market Hardware for Irregular, Low-Locality Applications

### 2012-07-23 doc/asplos13/intro.tex
patched up citations

Luckily, many of the important irregular applications (e.g., graph
processing, our focus in this paper) naturally offer large amounts of
concurrency. This immediately suggests taking advantage of concurrency
to tolerate the latency of data movement. The fully custom Tera
MTA-2~\cite{tera} \TODO{this is mta-1 citation} system is a classic example of supporting irregular
applications by using concurrency to hide latencies. It had a large
distributed shared memory with no caches.  On every clock cycle, each
processor would execute a ready instruction chosen from one of its 128
hardware thread contexts, a sufficient number to fully hide memory
access latency.  The network was designed with a single word injection
rate that matched the processor clock frequency and sufficient bandwidth
to sustain a reference from every processor on every clock cycle.
Unfortunately, while an excellent match to extremely irregular
applications, the MTA was not cost-effective on applications that could
exploit locality and had very poor single-thread performance, making
it a commercial failure. The Cray XMT approximates the Tera MTA-2,
substantially reducing its cost but not overcoming its narrow range of
applicability.

@@

In this paper we introduce Grappa, a software runtime system that allows
a commodity x86 distributed-memory HPC cluster to be programmed as if it
were a single large shared-memory machine and provides scalable
performance for irregular applications. Grappa is designed to smooth
over some of the performance discontinuities in commodity hardware,
giving good performance when there is little locality to be exploited
while allowing the programmer to exploit it when it is available. Grappa
leverages as much freely available and commodity infrastructure as
possible. We use unmodified Linux for the operating system. An off the
shelf user-mode infiniband device driver stack~\cite{Melonox?} is used
for network interfacing. MPI is used for process setup and
tear down. GASnet~\cite{gasnet:spec} is used as the underlying mechanism for
remote memory reads and writes using active message invocations. To this
commodity hardware and software mix Grappa adds three main software
components: (1) a \emph{lightweight tasking} layer that supports a
context switch in as few as 38ns and distributed load balancing; (2) a
\emph{distributed shared memory} layer that supports customary
operations such as \emph{read and write as well as remote operations}
such as fetch-and-add~\cite{fetchandadd}; and (3) a \emph{message
aggregation} layer that combines short messages to mitigate the
aforementioned problem that commodity networks are designed to achieve
peak bandwidth only on large packet sizes, yet irregular applications
tend to fetch only a handful of bytes at a time.

>>>>

Luckily, many of the important irregular applications (e.g., graph
processing, our focus in this paper) naturally offer large amounts of
concurrency. This immediately suggests taking advantage of concurrency
to tolerate the latency of data movement. The fully custom Tera
MTA-2~\cite{tera:mta1} system is a classic example of supporting irregular
applications by using concurrency to hide latencies. It had a large
distributed shared memory with no caches.  On every clock cycle, each
processor would execute a ready instruction chosen from one of its 128
hardware thread contexts, a sufficient number to fully hide memory
access latency.  The network was designed with a single word injection
rate that matched the processor clock frequency and sufficient bandwidth
to sustain a reference from every processor on every clock cycle.
Unfortunately, while an excellent match to extremely irregular
applications, the MTA was not cost-effective on applications that could
exploit locality and had very poor single-thread performance, making
it a commercial failure. The Cray XMT approximates the Tera MTA-2,
substantially reducing its cost but not overcoming its narrow range of
applicability.

@@

In this paper we introduce Grappa, a software runtime system that allows
a commodity x86 distributed-memory HPC cluster to be programmed as if it
were a single large shared-memory machine and provides scalable
performance for irregular applications. Grappa is designed to smooth
over some of the performance discontinuities in commodity hardware,
giving good performance when there is little locality to be exploited
while allowing the programmer to exploit it when it is available. Grappa
leverages as much freely available and commodity infrastructure as
possible. We use unmodified Linux for the operating system. An off the
shelf user-mode infiniband device driver stack~\cite{OFED} is used
for network interfacing. MPI is used for process setup and
tear down. GASnet~\cite{gasnet:spec} is used as the underlying mechanism for
remote memory reads and writes using active message invocations. To this
commodity hardware and software mix Grappa adds three main software
components: (1) a \emph{lightweight tasking} layer that supports a
context switch in as few as 38ns and distributed load balancing; (2) a
\emph{distributed shared memory} layer that supports customary
operations such as \emph{read and write as well as remote operations}
such as fetch-and-add~\cite{fetchandadd}; and (3) a \emph{message
aggregation} layer that combines short messages to mitigate the
aforementioned problem that commodity networks are designed to achieve
peak bandwidth only on large packet sizes, yet irregular applications
tend to fetch only a handful of bytes at a time.
